Queue Aware Precoder Design for Space Frequency Resource Allocation

### 2014-07-03 Report/joint_wsrm_formulation.tex
Update at home

The \ac{WSRM} technique is predominantly used to design the precoders for the \ac{MIMO} \ac{BC} channels to maximize the weighted sum throughput in the downlink direction. The weights are used to provide different priorities to the users based on the respective \ac{QoS} criterion. Since the objective of this work is concerned with the reduction of the number of queued packets, we use the total number of the queued packets for each user as the respective weights. Using the size of the queued packets as the corresponding weights is obtained from the well known algorithm, called the \textit{backpreassure algorithm}, which is used to decide the routing decisions for each node in the network to efficiently send the packets to the desired destination with minimal packet drops and delay. The backpreassure algorithm is the outcome of minimizing the Lyapunov drift \cite{neely2010stochastic} conditioned over the current queue states \me{\mbf{Q}(i - 1)}, where \me{\mbf{Q}} is the vector formed by stacking the number of queued packets of each user in the system. Assuming the queue length grows according to \eqref{eqn-2a} for all the users, the Lyapunov drift conditioned over the current queue state is given by
\begin{equation} \label{eqn-3.1.0}
\Delta(i) = \frac{1}{2} \, \sum_{k \in \mc{U}} \, \mathrm{E} \Big{[}  \left ( Q_k(i)^2 - Q_k(i-1)^2 \right ) \vert \, \mbf{Q}(i - 1) \Big{]},
\end{equation}
where the expectation is taken over all possible control\footnote{transmission rates} decisions made in accordance to the channel under consideration to minimize the total backlogged packets at a given instant. Now, by substituting \eqref{eqn-2a} in \eqref{eqn-3.1.0}, the resulting expression is given by
\begin{equation}
\Delta(i) \leq \mathrm{E} \Big{[} \mbf{Q}(i-1) \left \lbrace \mbfa{\lambda}(i) - \mbf{t}(i) \right \rbrace | \mbf{Q}(i-1) \Big{]} + \frac{1}{2} \,\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \frac{1}{2} \,\mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ],
\label{drift-exp}
\end{equation}
where \me{\mbfa{\lambda}} and \me{\mbf{t}} are the vectors formed by stacking the arrival and transmission of all users in the system and the inequality provides the upper bound\footnote{\me{\left ( \max{\Big [ \mbf{Q}(i-1) - \mbf{t}(i),0 \Big ] } + \mbfa{\lambda}(i) \right )^2 \leq \mbf{Q}^2(i-1) + \mbf{t}^2(i) + \mbfa{\lambda}^2(i) + 2 \, \mbf{Q}(i-1) \left ( \mbfa{\lambda}(i) - \mbf{t}(i) \right ) }} for the Lyapunov drift.

>>>>

The \ac{WSRM} technique is predominantly used to design the precoders for the \ac{MIMO} \ac{BC} scheme to maximize the weighted sum throughput for all users. The weights provide different priorities to the users based on their respective \ac{QoS} requirements. Since the objective of the current work is to reduce the total number of backlogged packets, the weights should be a function of the user queue sizes.

the reduction of the number of queued packets, we use the total number of the queued packets for each user as the respective weights. Using the size of the queued packets as the corresponding weights is obtained from the well known algorithm, called the \textit{backpreassure algorithm}, which is used to decide the routing decisions for each node in the network to efficiently send the packets to the desired destination with minimal packet drops and delay. The backpreassure algorithm is the outcome of minimizing the Lyapunov drift \cite{neely2010stochastic} conditioned over the current queue states \me{\mbf{Q}(i - 1)}, where \me{\mbf{Q}} is the vector formed by stacking the number of queued packets of each user in the system. Assuming the queue length grows according to \eqref{eqn-2a} for all the users, the Lyapunov drift conditioned over the current queue state is given by
\begin{equation} \label{eqn-3.1.0}
\Delta(i) = \frac{1}{2} \, \sum_{k \in \mc{U}} \, \mathrm{E} \Big{[}  \left ( Q_k(i)^2 - Q_k(i-1)^2 \right ) \vert \, \mbf{Q}(i - 1) \Big{]},
\end{equation}
where the expectation is taken over all possible control\footnote{transmission rates} decisions made in accordance to the channel under consideration to minimize the total backlogged packets at a given instant. Now, by substituting \eqref{eqn-2a} in \eqref{eqn-3.1.0}, the resulting expression is given by
\begin{equation}
\Delta(i) \leq \mathrm{E} \Big{[} \mbf{Q}(i-1) \left \lbrace \mbfa{\lambda}(i) - \mbf{t}(i) \right \rbrace | \mbf{Q}(i-1) \Big{]} + \frac{1}{2} \,\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \frac{1}{2} \,\mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ],
\label{drift-exp}
\end{equation}
where \me{\mbfa{\lambda}} and \me{\mbf{t}} are the vectors formed by stacking the arrival and transmission of all users in the system and the inequality provides the upper bound\footnote{\me{\left ( \max{\Big [ \mbf{Q}(i-1) - \mbf{t}(i),0 \Big ] } + \mbfa{\lambda}(i) \right )^2 \leq \mbf{Q}^2(i-1) + \mbf{t}^2(i) + \mbfa{\lambda}^2(i) + 2 \, \mbf{Q}(i-1) \left ( \mbfa{\lambda}(i) - \mbf{t}(i) \right ) }} for the Lyapunov drift.

### 2014-07-03 Report/problem_formulation_intro.tex
Update at home

In this paper, we address the problem of resource allocation across the \ac{MIMO} \ac{OFDM} framework, which are shared by the users served by multiple \acp{BS}. In order to minimize the queued packets, the precoders are to be designed in a way to overcome the problem of allocating the resources more than required. In order to satisfy this constraint, we use the queue deviation as the utility function as compared to the traditional \ac{WSRM} objective discussed extensively in the literature \cite{christensen2008weighted,wmmse_shi}. For practical and tractability reasons, we impose a constraint that the maximum number of transmitted bits for the user \me{k} is limited by the packets available at the transmitter. As a result, the number of backlogged packets \me{v_k} remaining in the system for the user \me{k} is given by
\begin{equation}
v_k =  Q_k - \sum_{n = 1}^N \sum_{l = 1}^{L} \log_2(1+\gamma_{l,k,n}) \geq 0 \fall k \in \mathcal{U}.
\label{eqn-4.2}
\end{equation}
The above positivity constraint need to be satisified by \me{v_k} to avoid the excessive allocation of the resources.

The precoder design problem can be given minimizing the absolute value of the deviation in \eqref{eqn-4.2} raised to the exponent \me{q} as \me{\text{minimize} \: \sum_{k \in \mc{U}} \, | v_k |^q}. The exponent \me{q} plays a vital role in the final allocation, which will be discussed later. Now, the problem of weighted queued packet minimization formulated as a $q$-norm minimization, is given by
\begin{IEEEeqnarray}{rCl}\label{eqn-3}
\underset{\substack{\mvec{m}{l,k,n},\\\mvec{W}{k,n}}}{\text{minimize}} &\quad& \|  \tilde{\mbf{v}}  \|_q\IEEEyessubnumber \\
\text{subject to} & \quad&\sum_{n = 1}^N \sum_{k \in \mathcal{U}_b} \text{tr} \, (\mvec{M}{k,n} \mvec{M}{k,n}^\herm) \leq P_{{\max}}, \fall b, \IEEEyessubnumber \label{eqn-4.3}
\end{IEEEeqnarray}
where \me{\tilde{v}_k \triangleq a_k^{1/{q}}v_k}, \me{a_k} is the weighting factor which is incorporated to control user priority based on their respective \ac{QoS}, \me{\gamma_{l,k,n}} is defined in \eqref{eq:SINR}, \me{\mbf{M}_{k,n} \triangleq [ \, \mvec{m}{1,k,n} \, \mvec{m}{2,k,n} \dotsc \mvec{m}{L,k,n} \,]} comprises the beamformers associated with the user \me{k} for \me{n^\mathrm{th}} sub-channel transmission, and \me{\mbf{W}_{k,n} \triangleq [ \, \mvec{w}{1,k,n} \, \mvec{w}{2,k,n} \dotsc \mvec{w}{L,k,n} \,]} stacks the receive beamformers respectively. It can be easily extended for user specific streams \me{L_k} instead of using the common \me{L} streams for all users. In \eqref{eqn-4.3}, we consider the sum power constraint for each \ac{BS} across all sub-channels. Before discussing the formulation in detail, we first discuss the existing algorithm to solve the current issue of minimizing the number of backlogged packets at the transmitter.

>>>>

In this paper, we address the problem of resource allocation across the \ac{MIMO} \ac{OFDM} framework, which are contended by the users served by multiple \acp{BS}. In order to minimize the number of queued packets, the precoders are designed to utilize the available space-frequency resources efficiently by limiting the over utilization of the same. In order to achieve this, we use the queue deviation utility as compared to the traditional \ac{WSRM} objective discussed extensively in the literature \cite{christensen2008weighted,wmmse_shi}. For practical and tractability reasons, we impose a constraint that the maximum number of transmitted bits for the user \me{k} is limited by the total backlogged packets available at the transmitter. As a result, the number of backlogged packets \me{v_k} remaining in the system for the user \me{k} is given by
\begin{equation}
v_k =  Q_k - \sum_{n = 1}^N \sum_{l = 1}^{L} \log_2(1+\gamma_{l,k,n}) \geq 0 \fall k \in \mathcal{U}.
\label{eqn-4.2}
\end{equation}
The above positivity constraint need to be satisified by \me{v_k} to avoid the excessive allocation of the resources.

The objective function for the precoder design problem is given by minimizing the absolute value of the queue deviation in \eqref{eqn-4.2} raised to the exponent \me{q} as \me{\text{minimize} \: \sum_{k \in \mc{U}} \, | v_k |^q}. With this objective, the problem of weighted queued packet minimization is given by the $q$-norm minimization as
\begin{IEEEeqnarray}{rCl}\label{eqn-3}
\underset{\substack{\mvec{M}{k,n},\gamma_{l,k,n}\\\mvec{W}{k,n}}}{\text{minimize}} &\quad& \|  \tilde{\mbf{v}}  \|_q\IEEEyessubnumber \\
\text{subject to} & \quad&\sum_{n = 1}^N \sum_{k \in \mathcal{U}_b} \text{tr} \, (\mvec{M}{k,n} \mvec{M}{k,n}^\herm) \leq P_{{\max}}, \fall b, \IEEEyessubnumber \label{eqn-4.3}
\end{IEEEeqnarray}
where \me{\tilde{v}_k \triangleq a_k^{1/{q}}v_k}, \me{a_k} is the weighting factor which is incorporated to control user priority based on their respective \ac{QoS}, \me{\gamma_{l,k,n}} is defined in \eqref{eq:SINR}, \me{\mbf{M}_{k,n} \triangleq [ \, \mvec{m}{1,k,n} \, \mvec{m}{2,k,n} \dotsc \mvec{m}{L,k,n} \,]} comprises the beamformers associated with the user \me{k} for \me{n^\mathrm{th}} sub-channel transmission, and \me{\mbf{W}_{k,n} \triangleq [ \, \mvec{w}{1,k,n} \, \mvec{w}{2,k,n} \dotsc \mvec{w}{L,k,n} \,]} stacks the receive beamformers respectively. It can be easily extended for user specific streams \me{L_k} instead of using the common \me{L} streams for all users. In \eqref{eqn-4.3}, we consider the sum power constraint for each \ac{BS} across all sub-channels. Before discussing the solutions in detail, we discuss the existing algorithm to solve the issue of minimizing the number of backlogged packets with additional constraints required by problem.

### 2014-07-03 Report/queue_minimizing_intro.tex
Updated ver_1
Update at home

In this section, we discuss the precoder design solutions for the problem stated in Section \ref{sec-3.2}. To begin with, we discuss the existing solution and its extension to solve the current problem of resource allocation over multiple sub-channels. Then, we provide two different approaches to solve the \ac{JSFRA} precoder design problem to minimize the total number of backlogged packets. At the end of the section, we discuss a low complex solution for the \ac{JSFRA} problem by limiting the precoder design over each sub-channel at a given instant.

In general, the precoder design for the \ac{MIMO} \ac{OFDM} problem is highly difficult due to the non-convex nature of the problem. In addition to that, the objective of minimizing the number of the queued packets over the spatial and the sub-channel dimensions adds further complexity to the existing problem. Since the scheduling of users in each sub-channel can be made by allocating zero transmit power over certain sub-channels, the solutions provided in the paper performs both precoder design and the scheduling of users in a joint manner.

>>>>

% To begin with, we discuss the existing solution and its extension to solve the problem of resource allocation over multiple sub-channels. Then, we provide two approaches to solve the \ac{JSFRA} precoder design problem to minimize the total number of backlogged packets. At the end, we discuss a low complex solution for the \ac{JSFRA} problem by limiting the precoder design over each sub-channel at a given instant.

In general, the precoder design for the \ac{MIMO} \ac{OFDM} problem is highly difficult due to the combinatorial and the non-convex nature of the problem. In addition to that, the objective of minimizing the number of the queued packets over the spatial and the sub-channel dimensions adds further complexity to the existing problem. Since the scheduling of users in each sub-channel can be made by allocating zero transmit power over certain sub-channels, the solutions provided in the paper performs both precoder design and the scheduling of users in a joint manner.

### 2014-07-03 Report/system_model.tex
Update at home

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} for transmission.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received data symbol corresponding to the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n^\mathrm{th}} at user $k$ is given by
\begin{equation}\label{eqn-1}
\hat{d}_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} with rank \me{L = \min(N_R,N_T)} is the channel between the \ac{BS} \me{b} and  user \me{k} on the sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. Assuming  \me{||\mvec{w}{l,k,n}||_{2}^{2} = 1} and independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}.
\end{equation}
With the infinite buffer model assumption, let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{i}^\mathrm{th}} instant for the user \me{k} depend on the fresh arrivals at the \me{{i}^\mathrm{th}} instant and the total number of backlogged packets \me{Q_k(i-1)} as given by
\begin{equation}
Q_k(i) = \Big [ Q_k(i-1) - t_k(i-1) \Big ]^+ + \lambda_k(i),
\label{eqn-2a}
\end{equation}
where \me{[x]^+ = \max{ \lbrace x,0  \rbrace}} and \me{t_k} denotes the transmission in bits for user \me{k}. For \ac{MIMO} \ac{OFDM} system,
\begin{equation}
t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i),
\end{equation}
where \me{t_{l,k,n} = \log_2(1 + \gamma_{l,k,n})} corresponds to the maximum rate achieved by Gaussian signaling on \me{l^\mathrm{th}} spatial stream and over \me{n^\mathrm{th}} sub-channel for the \ac{SINR} of \me{\gamma_{l,k,n}}. Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.

>>>>

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} for transmission.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received data symbol corresponding to the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n^\mathrm{th}} at user $k$ is given by
\begin{equation}\label{eqn-1}
\hat{d}_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} with rank \me{L = \min(N_R,N_T)} is the channel between the \ac{BS} \me{b} and  user \me{k} on the sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. Assuming  \me{||\mvec{w}{l,k,n}||_{2}^{2} = 1} and independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}.
\end{equation}
With the infinite buffer model assumption, let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{(i+1)}^\mathrm{th}} instant for the user \me{k}, represented by \me{Q_k(i+1)}, follows
\begin{equation}
Q_k(i+1) = \Big [ Q_k(i) - t_k(i) \Big ]^+ + \lambda_k(i),
\label{eqn-2a}
\end{equation}
where \me{[x]^+ \equiv \max{ \lbrace x,0  \rbrace}} and \me{t_k} denotes the transmission in bits for user \me{k}. For a \ac{MIMO} \ac{OFDM} system,
\begin{equation}
t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i),
\end{equation}
where \me{t_{l,k,n}} denotes the transmitted bits over \me{\ith{l}} spatial stream on the \me{\ith{n}} sub-channel. The maximum rate achieved over the \me{(l,n)} space-frequency resource is given by \me{\log_2(1 + \gamma_{l,k,n})}\footnote{assuming Gaussian signaling} for a given \ac{SINR} of \me{\gamma_{l,k,n}}. Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.
